### 2017-08-24 paper/dispRity_paper.tex
Updated paper

\section{Description}
In brief, the package intakes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows to analyse the resulting \disp object (see Supplementary @@@ for details on the object content).
Some additional functions allow to modify the multidimensional space such as dividing it per group or through time or/and bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
See below for a more detailed description of features.

@@

These divisions can readily be created in the package using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
The first takes as argument the matrix defining the multidimensional space and then a list of elements to group into different sub-samples.
The second also takes a multidimensional space and an additional argument on the age of the elements and then sorts them per time categories based on the following methods:

\begin{itemize}
    \item Discrete time subsamples: where the elements are grouped by age in time bins with corresponding ages.
    \item Continuous time subsamples: where the elements are defined as being the elements present at specific points in time (see @@@).
\end{itemize}

When using the \texttt{time.subsamples} function, it is also possible to provide a list of first and last occurence data for taxa spanning across specific time periods.
This allows species with a longer timespan to be present in multiple time sub-samples.

The continuous time sub-samples method works by using a phylogeny and looking at which taxa are present at any specific point in time.
This method thus requires a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.

The slicing works by selecting any tips (and nodes if \texttt{inc.nodes = TRUE}) at the specific time slice.
When a slice occurs not on a tip or a node, four methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
The models are:
\begin{itemize}
    \item ``acctran'' (accelerated transformation) where the data chosen along the branch is always the one of the descendant
    \item ``deltran'' (delayed transformation) where the data chosen along the branch is always the one of the ancestor
    \item ``punctuated'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor
    \item ``gradual'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length
\end{itemize}

\subsection{Bootstrapping and rarefying}
When measuring disparity, it is important to take sampling into account: for example, observed disparity in the whole multidimensional space can simply due to factors like the specific selection of elements (both their number or the presence of outliers).
To take these sources of error into account, it is common practice to bootstrap the data (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different sub-samples).

\subsubsection{Bootstraps}
In disparity analysis, if disparity is defined as a dimension-level 1 metric (i.e. a metric summarising the multidimensional space into a single value), it can be useful to bootstrap this value in order to have a distribution on which to perform statistical analysis.
This can be achieved by using the \texttt{boot.matrix} function pseudo-replicates the multidimensional space following these two algorithms:

\begin{itemize}
    \item ``full'' algorithm: where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data)
    \item ``single'' algorithm: where only one random element is replaced by one other random element for each pseudo-replicate
\end{itemize}

\noindent Note that the ``single'' algorithm is somewhat akin to jackknife re-sampling where one element is duplicated rather than removed.

In the case of disparity being a dimension-level 1 metric, it is important to note that the distribution resulting from a bootstrap will be a pseudo-replicated one and thus might not be suitable for some specific parametric statistical tests.
Conversely, it is possible to use a dimension-level 2 metric on a bootstrapped multidimensional space to avoid this problem: the disparity metric is then measured on a pseudo-replicated multidimensional space rather than the metric itself being pseudo-replicated.

\subsubsection{Rarefactions}
The same \texttt{boot.matrix} function also allows to rarefy the multidimensional space or its sub-samples.
Rarefaction can be useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of sub-samples wit the same number of elements in each.
In practice, rarefaction allows users to set a limit to the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).

Additionally, this argument, is set to \texttt{TRUE} also allows to fully rarefy the sub-samples (i.e. removing $x$ elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).

\subsection{Interpreting results}
Once disparity has been calculated (without or without sub-sampling, bootstrapping and rarefaction), the resulted \disp object can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity}.
Finally, the results can be analysed using the \texttt{test.dispRity} function.

\subsubsection{Summary and plot}
The summary and plot functions allow to summarise the contents of the \disp objects.
These functions take as common arguments which central tendency (e.g. if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).

Additionally, the \texttt{plot.dispRity} function allows to graphically represent the summarised results using different representations:

\begin{itemize}
    \item ``continuous'' for displaying continuous disparity curves (in the case of disparity through time analysis for example)
    \item ``box'', ``lines'', or ``polygons'' to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.
\end{itemize}

Additional arguments can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped), \texttt{add} to add the plot in the current quartz window, or \texttt{rarefaction} to only plot the disparity for a specific number of elements and of course, any other normal graphical options can also be passed to this function (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...).
Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.

\subsubsection{testing hypothesis}
The \texttt{test.dispRity} function allows to test hypothesis on the disparity data (e.g. by comparing the different sub-samples between each other).
Similarly to the \disp function described above, this one can take any test defined by the user.
The \texttt{comparison} arguments allows to indicate in which order (if any) should the tests be applied:

\begin{itemize}
    \item ``pairwise'' for applying pairwise comparisons of all the subsamples (default).
    \item ``referential'' for comparing the first subsample to all the others.
    \item ``sequential'' for comparing each subsample sequentially (e.g. first against second, second against third, etc.).
    \item ``all'' for comparing all the subsamples simultaneously to the data (i.e. \texttt{disparity $\mathtt{\sim}$ subsamples}). This argument can be used for \texttt{lm} or \texttt{glm} type tests.
    \item or any list of pairs (or more) of number of sub-samples to compare (for example, for comparing the first to the sixth subsample and the fifth to the second: \texttt{comparison = list(c(1, 6), c(5, 2))})
\end{itemize}

Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff}][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).

This function also allows additionally argument such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust eventual p-values when multiple parametric tests are ran.

\section{Examples}
Here I provide to classic example of disparity analysis using some of the functions described above.
The data to reproduce these examples is available within the package and thus don't need to be downloaded:

@@

\begin{itemize}
    \item \texttt{BeckLee\_mat50}: an ordinated matrix for 50 mammals based on the distance between discrete morphological characters.
    \item \texttt{BeckLee\_mat99}: an ordinated matrix for the same 50 mammals as in \texttt{BeckLee\_mat50} but also containing the reconstruction of their 49 ancestors.
    \item \texttt{BeckLee\_tree}: a chronogram with the 50 mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
    \item \texttt{BeckLee\_ages}: the first and last occurrence data for 14 of the mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
    \item \texttt{disparity}: a pre-made \disp object based on the data above.
\end{itemize}

\subsection{Typical disparity per groups analysis}

As an example, one can define their multidimensional space as being the morphospace of an ordination of the distance between discrete morphological characters for 50 mammals species \citep[from][]{beckancient2014}.
Then, one can define disparity as the sum of the variances on each dimensions \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e. a large sum of variances approximates a large morphospace).
One can then be interested in whether the approximation of this volume is different in the living mammals and their descendants (crown) or the fossil mammals with no living descendants (stem).
These two groups can be defined as follow (where the integers are the row number of the species in the matrix):

\texttt{> mammal\_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))}

@@

\texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50, group = mammal\_groups, metric = c(sum, variances))}

@@

\texttt{> subsamples <- custom.subsamples(data = BeckLee\_mat50, group = mammal\_groups)}\\
\texttt{> bootstraps <- boot.matrix(subsamples)}\\
\texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\

\noindent Which allows a finer tuning of each optional argument in each of the three functions \texttt{custom.subsamples}, \texttt{boot.matrix} and \disp.
The three arguments here are pretty self explanatory: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.

This function returns a \disp object that does not directly gives us the results (the disparity in each time bin) but rather summarises the content of the \disp object.

\noindent \texttt{ ---- dispRity object ---- }\\
\noindent \texttt{2 customised subsamples for 50 elements with 48 dimensions:}\\
\noindent \texttt{    crown, stem.}\\
\noindent \texttt{Data was bootstrapped 100 times (method:"full").}\\
\noindent \texttt{Disparity was calculated as: c(sum, variances).}\\

\bigskip
As indicated, the \disp object contains 2 customised subsamples (called ``crown'' and ``stem'') from a morphospace made of 48 dimensions and 50 elements.
The \disp object also displays information on the number and method of bootstrap replicates as well as how disparity was measured.
To visualise the actual disparity values, one can use the \texttt{summary} or/and \texttt{plot} options (Table \ref{Tab:summary_group} and Fig. \ref{Fig:plot_group}):

@@

\texttt{> plot(disparity)}

@@

\texttt{> test.dispRity(disparity, test = wilcox.test)}

@@

Another type of disparity analysis can be disparity through time analysis.
Using the same definition of the multidimensional space and disparity as in the previous example, one can be interested in how disparity changed through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).

\texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}

@@

\texttt{> disparity <- dispRity.through.time(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}

@@

\texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins)}\\
\texttt{> bootstraps <- boot.matrix(subsamples)}\\
\texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\

\noindent Which allows a finer tuning of each optional argument in each of the three functions \texttt{time.subsamples}, \texttt{boot.matrix} and \disp.
The arguments \texttt{data = BeckLee\_mat50} and \texttt{metric = c(sum, variances)} are the same as in the example above.
However, in this type of analysis, one also needs to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins and the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element.

@@

% latex table generated in R 3.4.1 by xtable 1.8-2 package
% Mon Aug 21 12:21:36 2017
\begin{table}[ht]
\centering
\begin{tabular}{rlrrrrrrr}
  \hline
 & subsamples & n & obs & bs.median & 2.5\% & 25\% & 75\% & 97.5\% \\ 
  \hline
1 & 100.5 - 66 &  15 & 1.67 & 1.55 & 1.40 & 1.51 & 1.58 & 1.65 \\ 
  2 & 66 - 56 &   9 & 1.88 & 1.69 & 1.43 & 1.63 & 1.77 & 1.83 \\ 
  3 & 56 - 33.9 &  13 & 1.96 & 1.83 & 1.62 & 1.77 & 1.86 & 1.90 \\ 
   \hline
\end{tabular}
\caption{Summarising a \disp object (disparity through time)}
\label{Tab:summary_time}
\end{table}

@@

From the plot and the summary, disparity seems to increase through time from the Cretaceous to the Eocene. Similarly to the example above, it is also possible to statistically test this hypothesis by running a simple ANOVA for example (Table \ref{Tab:anova}):

\texttt{> anova(test.dispRity(disparity, test = lm, comparisons = "all"))}

@@

\section{Additional informations}
\subsection{Manuals and vignette}
Supplementary information about the package is readily available on the project page \url{https://TGuillerme/dispRity}.
Additionally, information concerning each functions can be found in R in the individual function manuals an a vignette with more detailed information is available online at \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
This vignette contains substantial information on the functionalities of the package not described in this succinct description.
It also contains a more detailed tutorial of a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.

\subsection{Data simulations}
This package also contains a series of simulation functionality for generating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).
This functions are based on a similar modular architecture as the \disp functions allowing users to provide their own distribution descriptions as function as parameters for the simulation.
For example \texttt{stats::rnorm} can be provided for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.
The discrete morphological data simulations are based on protocols from \cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}. 
The multidimensional space simulations are based on the methods from \cite{diaz2016global}.
Both functionalities are described in more details in the package vignette.

\subsection{Further directions}
Further functionalities and improvement on current functionalities will be implemented in future versions.
In fact, optimised architecture and C implementations are being currently developed for faster calculations and lower memory footprints for most computational heavy functions (e.g. \disp or \texttt{time.subsamples}).
New functionalities currently being developed will include time sequential linear models (for accurately measuring directionality in disparity changes between time slices) or mode of disparity evolution testing (allowing to fit different models such as Brownian Motion or Ornstein-Uhlenbeck to disparity curves).

@@

%Editors:
%Sam Price, Daniele Silvestro

%Reviewers
%Graeme Lloyd, Dave Bapst, Emma Sherratt

>>>>

\section{Description}
In brief, the package intakes a \texttt{matrix} object (the multidimensional space), calculates some aspect of it (the disparity metric) and allows to analyse the resulting \disp object (see Manuals and Vignettes below).
Some additional functions allow to modify the multidimensional space such as dividing it per group or through time or/and bootstrapping it (see Fig. \ref{Fig:workflow} and Table \ref{Tab:main_fun}).
See below for a more detailed description of features.

@@

This can be done using the \texttt{custom.subsamples} or \texttt{time.subsamples} functions.
The first takes as argument the matrix defining the multidimensional space and a list of elements to group into different sub-samples.
The second also takes a matrix and arguments giving the age of the taxa (a phylogeny and/or a table containing the first and last occurence data for all/some elements - this allows species with a longer timespan to be present in multiple time sub-samples for example) and which subsamples to create: (1) discrete time subsamples (or time-binning) or (2) continuous time subsamples (or time-slicing).

The time-bining method groups elements by specific age range.
The time slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
This method thus requires the nodes to be part of the multidimensional space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
When a slice occurs not on a tip or a node, four methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
\begin{itemize}
    \item ``acctran'' (accelerated transformation) where the data chosen along the branch is always the one of the descendant
    \item ``deltran'' (delayed transformation) where the data chosen along the branch is always the one of the ancestor
    \item ``punctuated'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor
    \item ``gradual'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length
\end{itemize}

\subsection{Bootstrapping and rarefying}
Disparity measurement can be influenced by sampling \citep{Butler2012}.
To take this source of bias into account, it is common practice to bootstrap the multidimensional space (to minimise the effect of outliers) or/and to rarefy the data (to minimise the effect of different sample sizes in different sub-samples).
Additionally, if disparity is defined as a dimension-level 1 metric, it can be useful to measure it on bootstrapped data in order to have a distribution on which to perform statistical analysis.

Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the multidimensional space following two algorithms (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data) and (2) the ``single'' one where only one random element is replaced for each pseudo-replicate (somewhat akin to jackknife resampling).

% In the case of disparity being a dimension-level 1 metric, it is important to note that the distribution resulting from a bootstrap will be a pseudo-replicated one and thus might not be suitable for some specific parametric statistical tests.
% Conversely, it is possible to use a dimension-level 2 metric on a bootstrapped multidimensional space to avoid this problem: the disparity metric is then measured on a pseudo-replicated multidimensional space rather than the metric itself being pseudo-replicated.

Similarly, rarefaction can be achieved through the same \texttt{boot.matrix} function.
Rarefaction can be useful when looking at the effect of reducing the number of elements on the results of an analysis or for allowing the comparisons of sub-samples wit the same number of elements in each.
In practice, rarefaction limits the number of elements to be drawn for each bootstrap replication: only $n-x$ elements are selected at each bootstrap replicate (where $x$ is the number of non-sampled elements).

\subsection{Interpreting results}
The functions above all generate a \disp object that can be summarised or plotted using the \texttt{S3} method functions \texttt{summary.dispRity} and \texttt{plot.dispRity} (provided disparity was calculated).
These results can then be analysed using the \texttt{test.dispRity} function for comparing sub-samples or testing hypothesis.

\subsubsection{Summary and plot}
These \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow to set which central tendency and which quantiles should be represented.
% as common arguments which central tendency (e.g. if central tendency is the mean and disparity was calculated as the sum of the variances for each dimensions, the central disparity tendency would be the mean of the sum of the variances) and the quantiles to represent (in the same example, if the quantile chosen is the 50th, these functions will also display 50\% of the sum of variances around the mean).
Additionally, the \texttt{plot.dispRity} function allows to graphically represent the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves (in the case of disparity through time analysis for example) and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for 
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...).
% Note that if the disparity has been calculated on fully rarefied data (using \texttt{boot.matrix(data, rarefaction = TRUE)} prior to calculating disparity), the rarefaction curves for the whole multidimensional space (or its sub-samples) can be plotted using \texttt{rarefaction = ``plot''}.

\subsubsection{testing hypothesis}
The \texttt{test.dispRity} function allows to test hypothesis on the disparity data (e.g. by comparing the different sub-samples between each other).
Similarly to the \disp function described above, this one can take any test defined by the user.
The \texttt{comparison} arguments allows to indicate in which order (if any) should the tests be applied to the subsamples: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subsample to all the others; (3) ``sequential'' for comparing subsamples sequentially (e.g. first against second, second against third, etc.); (4) `all'' for comparing all the subsamples simultaneously (i.e. \texttt{disparity $\mathtt{\sim}$ subsamples} - this can be used for \texttt{lm} or \texttt{glm}) or (5) any list of pairs (or more) of sub-samples to compare,

Some tests are implemented within the package such as the Bhattacharrya Coefficient \citep[\texttt{bhatt.coeff}][]{Bhattacharyya,GuillermeCooper} or a permutation test based on null hypothesised multidimensional space following \cite{diaz2016global} (\texttt{null.test}).
Additionally, this function also allows additionally argument such as \texttt{rarefaction} (as described above) or \texttt{correction} to adjust eventual p-values when multiple parametric tests are ran.

\section{Examples}
For brevity reason, the following example is based on a classical palaeobiology morphological disparity analysis.
Note the more examples are available in the package manual \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.

@@

\begin{itemize}
    \item \texttt{BeckLee\_mat50}: an ordinated matrix for 50 mammals based on the distance between discrete morphological characters.
    \item \texttt{BeckLee\_mat99}: the same matrix \texttt{BeckLee\_mat50} but also containing the reconstruction of their 49 ancestors.
    \item \texttt{BeckLee\_tree}: a chronogram with the 50 mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
    \item \texttt{BeckLee\_ages}: the first and last occurrence data for 14 of the mammals species present in \texttt{BeckLee\_mat50} and \texttt{BeckLee\_mat99}.
    \item \texttt{disparity}: a pre-made \disp object based on the data above.
\end{itemize}

In this example, the multidimensional space is thus defined as a morphospace: the ordination of the distance between discrete morphological characters for 50 mammals species \citep[from][]{beckancient2014}.
Additionally, we can define disparity as the sum of the variances on each dimensions \citep{Wills1994} that will represent an aspect of the the volume of the morphospace (i.e. a large sum of variances approximates a large morphospace and vice versa).

\subsection{Typical disparity per groups analysis}
Here, we can be interested in whether the approximation of the volume within the morphospace (the disparity as defined above) is different in the living mammals and their descendants (crown) or the fossil mammals with no living descendants (stem).
These two groups can be defined as follow (where the integers are the row number of the species in the matrix):

\noindent \texttt{> mammal\_groups <- list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44))}

@@

\noindent \texttt{> disparity <- dispRity.per.group(data = BeckLee\_mat50, group = mammal\_groups, metric = c(sum, variances))}

@@

\noindent \texttt{> subsamples <- custom.subsamples(data = BeckLee\_mat50, group = mammal\_groups)}\\
\noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\

\noindent Which allows a finer tuning of the optional argument in each functions.
The three arguments here are defined as follow: \texttt{data = BeckLee\_mat50} is our definition of the multidimensional space, \texttt{group = mammal\_groups} indicates which mammals belong to which group and \texttt{metric = c(sum, variances)} is our definition of disparity.

This function returns a \disp object that summarises the disparity analysis:

\noindent \texttt{> disparity}
\noindent \texttt{ ---- dispRity object ---- }\\
\noindent \texttt{2 customised subsamples for 50 elements with 48 dimensions:}\\
\noindent \texttt{    crown, stem.}\\
\noindent \texttt{Data was bootstrapped 100 times (method:"full").}\\
\noindent \texttt{Disparity was calculated as: c(sum, variances).}\\

\bigskip
As indicated, the \disp object contains 2 customised subsamples (called ``crown'' and ``stem'') from a morphospace made of 50 elements for 48 dimensions.
The \disp object also displays information on the number and method of the bootstrap replicates as well as the definition of disparity.
To visualise the actual disparity values, one can use the \texttt{summary} or/and \texttt{plot} options (Table \ref{Tab:summary_group} and Fig. \ref{Fig:plot_group}):

\noindent \texttt{> summary(disparity)}

@@

\noindent \texttt{> plot(disparity)}

@@

\noindent \texttt{> test.dispRity(disparity, test = wilcox.test)}

@@

Another type of disparity analysis can be a disparity through time analysis.
Using the same definition of the multidimensional space and disparity as in the previous example, one can be interested in how disparity changes through time between the Late Cretaceous (100.5 to 66.0 million years ago - Mya), the Paleocene (66.0 to 56.0 Mya) and the Eocene (56.0 to 33.9 Mya).

\noindent \texttt{> time\_bins <- c(100.5, 66, 56, 33.9)}

@@

\noindent \texttt{> disparity <- dispRity.through.time(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, metric = c(sum, variances))}

@@

\noindent \texttt{> subsamples <- time.subsamples(data = BeckLee\_mat50, tree = BeckLee\_tree, time = time\_bins, method = 'discrete')}\\
\noindent \texttt{> bootstraps <- boot.matrix(subsamples)}\\
\noindent \texttt{> disparity <- dispRity(bootstraps, metric = c(sum, variances))}\\

\noindent The arguments \texttt{data = BeckLee\_mat50} and \texttt{metric = c(sum, variances)} are the same as in the example above.
However, in this type of analysis, we also needs to have additional arguments: the \texttt{time = time\_bins} indicates the boundaries of the different time bins, the \texttt{tree = BeckLee\_tree} argument provides information on the age of each element and \texttt{method = 'discrete'} indicates that the data is time-binned.

@@

\begin{table}[ht]
\centering
\begin{tabular}{rlrrrrrrr}
  \hline
 & subsamples & n & obs & bs.median & 2.5\% & 25\% & 75\% & 97.5\% \\ 
  \hline
1 & 100.5 - 66 &  15 & 1.67 & 1.55 & 1.40 & 1.51 & 1.58 & 1.65 \\ 
  2 & 66 - 56 &   9 & 1.88 & 1.69 & 1.43 & 1.63 & 1.77 & 1.83 \\ 
  3 & 56 - 33.9 &  13 & 1.96 & 1.83 & 1.62 & 1.77 & 1.86 & 1.90 \\ 
   \hline
\end{tabular}
\caption{Summarising a \disp object (disparity through time)}
\label{Tab:summary_time}
\end{table}

@@

From the plot and the summary, disparity seems to change (increase) through time from the Cretaceous to the Eocene.
Similarly to the example above, it is also possible to statistically test this hypothesis by running an ANOVA for example (Table \ref{Tab:anova}):

\texttt{> anova(test.dispRity(disparity, test = lm, comparisons = 'all'))}

@@

\section{Additional informations}
\subsection{Manuals and vignette}
Supplementary information about the package is available on the project page \url{https://TGuillerme/dispRity}.
Information concerning each functions can be found in R in the individual function manuals and a package manual with more detailed information is available online at \url{https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/index.html}.
This manual contains substantially more information and detailed example including a tutorial for a ``classic'' disparity analysis in palaeobiology as well as an introduction to the use of this package in ecology or other disciplines.

\subsection{Data simulations}
This package also contains a series of simulation functionality for generating random discrete morphological matrices (\texttt{sim.morpho}) or random multidimensional spaces (\texttt{space.maker}).
These functions are based on a similar modular architecture as the \disp functions allowing users to provide their own distribution descriptions as parameters for the simulation.
For example, \texttt{stats::rnorm} can be provided as an argument for drawing normal characters rates with \texttt{sim.morpho} or normally distributed multidimensional space with \texttt{space.maker}.
The discrete morphological data simulations are based on protocols from \cite{GuillermeCooper}, \cite{OReilly20160081} and \cite{puttick2017uncertain}. 
The multidimensional space simulations are based on the methods from \cite{diaz2016global}.
Both functionalities are described in more details in the package manual.

% \subsection{Further directions} TG: maybe this section is useless
% Further functionalities and improvement on current functionalities will be implemented in future versions.
% In fact, optimised architecture and a C implementations are being currently developed for faster calculations and lower memory footprints for most computational heavy functions (e.g. \disp or \texttt{time.subsamples}).
% New functionalities currently being developed will include time sequential linear models (for accurately measuring directionality in disparity changes between time slices) or mode of disparity evolution testing (allowing to fit different models such as Brownian Motion or Ornstein-Uhlenbeck to disparity curves).

@@
